Building an SSI Cluter with GNU/Hurd

### 2019-05-12 libpager/paper.tex
working on paper.tex

@@

CAP THEOREM

The CAP theorem states that we can't have all three of: consistency,
availability, partitioning.  Right now, we sacrifice availability - if
the network partitions, libpager can deadlock.  If we introduce a
timeout on lock requests, then we're sacrificing consistency.

@@

In addition to opaque data (generally, but not necessarily interpreted
as remote procedure calls), Mach messages can contain two more
sophisticated structures.  The first are port rights.

A Mach message can transport send or receive rights for Mach ports.
This is basically the only way in which a process can obtain Mach
ports since, much like POSIX file descriptors, port numbers are local
to a process and interpreted by the kernel, so there is no way that a
process can fabricate another process's port number, just as a POSIX
process can not fabricate another process's file descriptor.

@@

Mach's memory management was specifically designed to support
multi-node clusters.  Mach's model of memory management is that memory
is managed by user-space processes, and system memory is a cache of
that memory.  Any user-space process that implements Mach's memory
management protocol can be mapped into a process's address space.
When the process attempts to access memory, a page fault is triggered
which causes Mach to request the page from the user-space process.
The faulting thread remains suspended until the page has been supplied
to the kernel.

@@

Of course, it can be inefficient to keep a copy of a page in user
space while its also in use by the kernel.  Pages can be flagged
``precious'', indicating to the kernel that they must be returned to
the user-space process and can not be simply discarded.

The flexibility of allowing user-space processes to implement the
memory management protocol does create unique failure and abuse
scenarios.  A program can implement the memory management protocol,
map it into a region of memory, transfer that memory to another
process, then trigger a fault by responding with
memory_object_data_error to any access attempt.  The other program, if
not careful, could fail with a SIGSEGV just for trying to read the
memory it received in a message!  Thus, in Hurd, user-space programs,
especially system servers, must exhibit the same kind of care when
receiving memory regions in Mach messages as Linux kernel authors are
used to considering when a kernel routine receives memory from a
user-space program.

@@

pager_read_page() is expected to allocate memory (unless it returns an
error) and pass it to libpager, which is responsible for deallocating
it (actually by passing it to the kernel).

pager_write_page() does no memory allocation or deallocation.  It
receives allocated memory and leaves it allocated, with libpager
responsible for dellocation (or passing it to the kernel).

pager_flush() and pager_flush_some() discard dirty pages!

pager_return() and pager_return_some() flush and return dirty (and precious) pages.

pager_sync() and pager_sync_some() return dirty (and precious) pages but don't flush.

@@

CONFIGURATION OPTIONS

- do we write a page back every time we handle it?
  perhaps always call write_page, and add a flag to write_page indicating that we still have a copy?
- how long do we let clients hold a page if other clients are waiting for it?
  currently: 0 (immediately send lock requests)
  could create deadlock if the kernel returns pages without making any forward progress
- how do we obtain pages when we can read them from multiple sources?
  only applies to READ access; local kernel should be preferred
- handle network partitions / failed nodes / malfunctioning kernels
  timeout while waiting for a client?
  CAP theorem - timeouts would sacrifice consistency to achieve availability

@@

Hurd's original \libpager had a fairly simple pagemap.  The pagemap was
an array of 16-bit shorts; each page had a single 16-bit entry, allocated
as follows:

@@

I used the clever copy constructor.  The pagemap is an
array of pointers to unique pagemap structures kept in a C++ {\tt std::set}
called {\tt pagemap_set}.  When working on a page, the main library
code copies a pagemap entry into a temporary structure ({\tt
  tmp_pagemap_entry}), modifies it, then assigns it back into the
pagemap using an {\tt operator=} that either finds a matching entry in
{\tt pagemap_set} and uses a pointer to it, or move-inserts {\tt
  tmp_pagemap_entry} into {\tt pagemap_set} and uses a pointer to it.

@@

   /* ERROR
    * the last error code returned from a backing store write operation
    * (hopefully KERN_SUCCESS)
    */
   kern_return_t error;
};

@@

   /* WRITEWAIT list
    * A linked list of pages waiting to be written out to disk
    */

@@

    std::set<outstanding_change_request> outstanding_change_requests;

   ERROR code - a kern_return_t, so we're not limited to a small number of errors

   /* Flags */
   PAGINGOUT - set this flag when we get a data return, clear it after the write is finished
   INVALID - set if the data on disk is invalid because we got an error writing the page back to the filesystem
      If some clients still have copies of the data, we may be able to operate for a while, then retry the write operation
}
\end{verbatim}

How to maintain the pagemap for a large number of clients?  Each page
has a pointer to a structure indicating which clients currently hold
copies of the page.  We anticipate that many pages (say, all the pages
in a single shared library) will have the same set of clients, so
there's one structure for each combination of clients, with multiple
pages pointing to it.  Pagemaps can be potentially large, so a pagemap
entry should be a single pointer.

@@

@@

BUG IN OLD CODE:
  m_o_data_supply is called in data-return.c with precious = 0, instead of notify_on_evict

BUG IN OLD CODE:
  in pager_shutdown(), pthread_mutex_unlock() is called after ports_destroy_right(), which
  might have free'd the struct pager

@@

Scenario A
- five clients have read access to page 50 and are on its ACCESSLIST (WAITLIST is empty)
- pager_flush is called on pages 0-100
- page 50 flushes on client A
- client A re-requests access to page 50, and it is supplied (ACCESSLIST and WAITLIST don't change)
- the flushes complete
- only client A now has access to page 50 and should be alone on its ACCESSLIST

@@

translator operations:
1. discard request
2. return request
3. sync request
4. complete a pending pager_read, pager_write, or pager_unlock (which one? there could be several)

1 2 3 can be synchronous or asynchronous

The test program should check to ensure that following conditions are met:

[ ] write access to a page is exclusive to a single client
[p] pages get written back in-order (though skips are possible)
[ ] if a page is modified, nobody ever sees an earlier version of it
[ ] all data requests are properly answered with a data supply or a data error (or data unavailable)
[ ] all data unlock requests are properly answered with a data lock
[ ] if a data lock requests a reply, make sure one is received

p = partially done

@@

\item the ability to request a copy of a unmodified, un-precious page from the kernel

@@

    (over the network, precious is heavy weight just to get notification that a page has been evicted)

@@

    this would allow sharing pagemaps between memory objects with the same usage pattern
    (i.e, kernels A and B both map files C and D)

>>>>

\usepackage{../slides/hurd}
\usepackage{tikz}
\usetikzlibrary{positioning, fit, backgrounds, arrows, trees, shapes}

@@

The CAP theorem states that a distributed system can't have all of
three things: consistency, availability, partitioning.  Right now, we
sacrifice availability -- if the network partitions, libpager can
deadlock.  If we introduce a timeout on lock requests, then we're
sacrificing consistency.

@@

Mach inter-process communications is primarily done using
variable-sized datagram messages that are transmitted reliably and
in-order to queues from which tasks can receive them.  In addition to
opaque data (generally, but not necessarily interpreted as remote
procedure calls), Mach messages can contain two more sophisticated
structures.  The first are port rights, which come in two basic
flavors.  Send rights give a task the ability to transmit messages to
a queue, and receive rights give a task the ability to receive
messages from a queue.  A Mach message can transport send or receive
rights for Mach ports.

\tikzstyle{message} = [draw, fill=red!20, text width=5em, text depth=3em, text centered,
  minimum width=5em, minimum height=2em, rounded corners]

\begin{figure}
\begin{center}
\begin{tikzpicture}[node distance = .5cm]

\task{Task 1}{at (0,0)}{br}{
% print two sendright nodes, one to fill the message1 box to the correct size,
% and another one to actually draw it, since message1 will overwrite the
% first sendright

\node (sendright) [draw, fill=orange] {Send 41};
\node (message1) [fit=(sendright), message] {Message 1};
\node (sendright) [draw, fill=orange] {Send 41};

\node (message3) at (0, -7em) [message] {Message 2};
}

\task{Task 2}{at (15 em,0)}{bl}{

\node (sendright2) [draw, fill=orange] {Send 9};
\node (message2) [fit=(sendright2), message] {Message 1};
\node (sendright2) [draw, fill=orange] {Send 9};

\node (message4) at (0, -7em) [message] {Message 2};
}

\node (port 41) [left of=Task 1 Port -10, anchor=south] {41};

\draw [thick, color=red, dotted, ->] (node cs:name=sendright, angle=-20) -- (port 41);

\node [left of=Task 1 Port 5, anchor=south] {27};
\node [right of=Task 2 Port 5, anchor=south] {52};
%% \only<1,2>{\draw [thick, -triangle 90] (Task 1 Port 5) -- (Task 2 Port 5);}

\draw [color=violet, thick, -triangle 90] (node cs:name=message1,angle=20) -- +(.3,0) |- (Task 1 Port 5) -- (Task 2 Port 5)
    -- +(.5,0) |- (node cs:name=message2, angle=160);

\node [right of=Task 2 Port -10, anchor=south] {9};

\draw [thick, triangle 90-] (Task 1 Port -10) -- (Task 2 Port -10);

\draw [color=violet, thick, triangle 90-] (node cs:name=message3,angle=20) -- +(.3,0) |- (Task 1 Port -10) -- (Task 2 Port -10)
    -- +(.5,0) |- (node cs:name=message4, angle=160);

\end{tikzpicture}
\end{center}
\caption{Transferring a send right}
\label{Transferring a send right}
\end{figure}

In Figure \ref{Transferring a send right}, Process 1 sends Message 1
to Process 2 containing a send right to the queue known to Process 1
as port 41, which could have been created by Process 1, or received
from another task.  The Mach kernel translates this into a send right
to a previously unused port 9 in Process 2's port number space.
Process 2 can now send a message to its port 9, which will be received
by Process 1 on port 41.

This is basically the only way in which a process can obtain Mach
ports since, much like POSIX file descriptors, port numbers are local
to a process and interpreted by the kernel, so there is no way that a
process can fabricate another process's port number, just as a POSIX
process can not fabricate another process's file descriptor.  A
process is provided a few standard ports when it is initialized, one
of which being a port with send rights that represents the root of the
file system.  By sending filename lookup requests to this port, the
process can traverse the entire file system tree, each node of which
is represented by a different port.  The hurd file system is thus a
hierarchy of ports, most of which recognize a standard set of RPCs
performing file operations.  The primary way for a program to
advertise its services in Hurd is for it to listen for messages on a
port that it attachs to a name in the file system.

@@

Mach's memory management was specifically designed to support
multi-node clusters.  Mach's model of memory management is that memory
is managed by user-space processes, and system memory is a cache of
that memory.  Any user-space process that implements Mach's memory
management protocol can be mapped into a process's address space.
When the process attempts to access memory, a page fault is triggered
which causes Mach to request the page from the user-space process
using a well documented protocol.  The faulting thread remains
suspended until the page has been supplied to the kernel.

@@

Of course, it can be inefficient to keep a copy of a page in user
space while its also in use by the kernel.  Pages can be flagged
``precious'', indicating to the kernel that they must be returned to
the user-space process and can not be simply discarded.  This frees
the user-space memory manager from needing to keep a copy of a page
that has been supplied to the kernel.  Otherwise, non-``precious''
pages can be freely discarded by the kernel when they are no longer
needed.

The flexibility of allowing user-space processes to implement the
memory management protocol does create unique failure and abuse
scenarios.  A program can implement the memory management protocol,
map it into a region of memory, transfer that memory to another
process, then trigger a fault by responding with
memory_object_data_error to any access attempt.  The other program, if
not careful, could fail with a SIGSEGV just for trying to read the
memory it received in a message!  Thus, in Hurd, user-space programs,
especially system servers, must exhibit the same kind of care when
receiving memory regions in Mach messages as Linux kernel routines
must consider when receiving memory from a user-space program.

@@

On one side, \libpager implements the Mach memory management protocol
interacting with the Mach kernel.  On the other side, \libpager
presents an API to the server program, which is responsible for
producing the pages, as follows:

\item {\tt pager_read_page()}

This routine is expected to allocate memory (unless it returns an
error) containing the requested data and pass it to \libpager, which
is responsible for deallocating it (typically by passing it to the
kernel).  A common use scenario is that {\tt pager_read_page()}
reads a page from disk and returns it.  {\tt pager_read_page()}
can be called from multiple threads to service multiple paging
requests, and must implement its own thread safety.

\item {\tt pager_write_page()}

This routine is expected to save a modified memory page just before
\libpager releases its last copy of the data.  It does no memory
allocation or deallocation.  It receives allocated memory and leaves
it allocated, with \libpager responsible for dellocation, or passing
it to the kernel, if a new request has arrived in the time required
for {\tt pager_write_page()} to complete

\item {\tt pager_unlock_page()}

\item {\tt pager_sync()} and {\tt pager_sync_some()}

This routine requests that copies of modified pages be obtained
and submitted to {\tt pager_write_page()}.  Doesn't otherwise
interfere with \libpager operation.

\item {\tt pager_flush()} and {\tt pager_flush_some()}

discard dirty pages!

\item {\tt pager_return()} and {\tt pager_return_some()}

 flush and return dirty (and precious) pages.

\item {\tt pager_get_error()}

@@

\section{Configuration Options}

Several implementation choices exist, and perhaps should be configuration options.

\item How long do we let clients hold a page if other clients are waiting for it?

  Currently we send a lock request immediately after supplying the
  page, which could create deadlock if the kernel returns pages
  without making any forward progress.  If processes on two different
  nodes are both trying to write into a page, we might spend all of
  our time ping-ponging the page and forth without anything actually
  getting done.

  Check Mach kernel source to see what happens in this case.  My guess
  is that if the process can be started immediately, it is, and this
  is only an issue if system load is high enough to prevent that.

\item Do we write a page back every time we handle it?

  Currently, we don't call {\tt pager_write_page()} until a page is
  completely freed or sync'ed, but this can cause modified data to sit
  indefinitely in RAM when we'd rather save it to disk periodically.

  Note that if the server wishes to periodically sync data to disk, it
  can call {\tt pager_sync()} on a timer thread.  The question is
  whether we want to more proactively call {\tt pager_write_page()}
  anytime we're handling a page that needs to be transferred between
  two clients.

\item How do we obtain pages when we can read them from multiple sources?

  Currently, we read it again with {\tt pager_read_page()}, but if the
  local kernel has a copy, it's probably faster to ask the kernel for
  a copy.  We don't have a good way of detecting which client is the
  local kernel, and the Mach memory management API doesn't provide a
  way to request a copy of an unmodified, non-``precious'' page
  from the kernel.

  A corner case is what happens when the read fails with an error.
  Currently, we propagate that error to everything waiting on the
  page, even though we could request a copy of the page from one
  of the clients that still has it.

\item How to handle network partitions / failed nodes / malfunctioning kernels

  Currently we deadlock if a client doesn't respond; perhaps we should
  timeout while waiting for a client?  How to pick the timeout value?

\item How to handle failed writes to the backing store?

  Do we keep the page around in memory to service future requests, and
  hope that maybe another write will succeed in the future, or discard
  the page?  Perhaps keep it for a time?  How to decide when to
  discard?  Maybe only keep it if EAGAIN is returned?

  If we do discard the data, how do we handle future requests for that
  page?  Currently, we return errors with no mechanism to recover.

  If we kept failing writes around indefinitely, that would provide a
  simple mechanism to implement a ramdisk -- the read operation always
  returns an empty page and the write operation always fails.

@@

All Mach memory management operations operate on memory pages, so
\libpager needs to maintain a table that tracks the state of each
page, called the pagemap.  Hurd's original \libpager had a fairly
simple pagemap.  The pagemap was an array of 16-bit shorts; each page
had a single 16-bit entry, allocated as follows:

@@

I used the clever copy constructor.  The pagemap is an array of
pointers to unique pagemap structures kept in a C++ {\tt std::set}
called {\tt pagemap_set}.  When working on a page, the main library
code copies a pagemap entry into a temporary structure ({\tt
  tmp_pagemap_entry}), modifies it, then assigns it back into the
pagemap using an {\tt operator=} that either finds a matching entry in
{\tt pagemap_set} and uses a pointer to it, or move-inserts {\tt
  tmp_pagemap_entry} into {\tt pagemap_set} and uses a pointer to it.
No attempt is (currently) made to free unused entries in {\tt
  pagemap_set}.

@@

   /* ERROR
    * the last error code returned from a backing store operation
    * (hopefully KERN_SUCCESS)
    */
   kern_return_t error;

   /* INVALID
    * indicates that the backing store data is invalid because we got an
    * error return from a write attempt
    *
    * error returns on read or unlock operations do not set "invalid"
    */
   bool invalid;
};

@@

   /* WRITEWAIT list
    * A linked list of pages waiting to be written to backing store
    */

@@

    std::set<outstanding_change_request> outstanding_change_requests;
}
\end{verbatim}

How to maintain the pagemap for a large number of clients?  Each page
has a pointer to a structure indicating which clients currently hold
copies of the page.  We anticipate that many pages (say, all the pages
in a single shared library) will exist in the same state, so there's
one structure for each combination of its entries, with multiple pages
pointing to it.  Pagemaps can be potentially large, so a pagemap entry
should be a single pointer.

@@

ERROR HANDLING

The three backing store callback functions can all return an error
code.

Errors when reading from backing store are fairly straightforward --
we got an error, so all we can do is propagate the error to the
clients.  Actually, if there are outstanding copies of the page we
tried to read, we could obtain a copy from another client, but we
don't do that currently.  Errors when writing are a bit more
complicated -- we've still got a copy of the page, do we keep it
around and hope that the write will complete successfully at some
point in the future?  We don't do that currently, either.  Errors
returned from an unlock operation should probably only affect the
unlock request.  Should they also affect future unlock operations on
this page, or should failing unlocks be retried?  Currently, a failing
unlock is propagated to everything waiting on the page (both read and
write requests); probably should only be sent to write requests.

Also, error returns on unlock requests are inconsistently recorded in
the pagemap.  The error return code in the pagemap is only really used
if a write operation failed, in which case the page is discarded and
future requests all return the error code.  An exception is if a
kernel kept a copy of the page whose write failed (i.e, there was a
sync request that triggered the write), in which case future
operations fail until something causes the kernel to send us the page
again, when we retry the write.

@@

@@

Scenario A
\begin{itemize}
\item five clients have read access to page 50 and are on its ACCESSLIST (WAITLIST is empty)
\item pager_flush is called on pages 0-100
\item page 50 flushes on client A
\item client A re-requests access to page 50, and it is supplied (ACCESSLIST and WAITLIST don't change)
\item the flushes complete
\item only client A now has access to page 50 and should be alone on its ACCESSLIST
\end{itemize}

@@

There are four possible translator operations:

\begin{enumerate}
\item discard request
\item return request
\item sync request
\item complete a pending pager_read, pager_write, or pager_unlock (which one? there could be several)
\end{enumerate}

1, 2, and 3 can be synchronous or asynchronous.

The test program should (but currently doesn't) check to ensure that
following conditions are met:

\begin{itemize}
\item write access to a page is exclusive to a single client
\item pages get written back in-order (though skips are possible) (partially done)
\item if a page is modified, nobody ever sees an earlier version of it
\item all data requests are properly answered with a data supply or a data error (or data unavailable)
\item all data unlock requests are properly answered with a data lock
\item if a data lock requests a reply, make sure one is received
\end{itemize}

@@

\item ability to request a copy of a unmodified, un-precious page from the kernel

@@

  Over the network, flagging a page ``precious'' is awfully heavy
  weight just to get notification that a page has been evicted, since
  a ``precious'' page will be sent back to the memory manager.
  Knowing that a page has been evicted may eliminate the need for a
  future lock request.

@@

  This would allow sharing pagemaps between memory objects with the same usage pattern
  (i.e, kernels A and B both map files C and D)

\item Add the ability for a kernel to respond to a lock request by
  sending the page directly to another kernel, rather than back to the
  memory manager.

  This would cut our network utilization roughly in half for pages
  dominated by contention between nodes.
